Omics-based Single Step Trait Prediction

### 2017-09-06 paper/manuscript.tex
fix latex warnings in manuscript

\renewcommand{\CorrespondingAddress}{%
    University of Hohenheim \\
    Institute of Plant Breeding, Seed Science and Population Genetics \\ 
    Fruwirthstr. 21 \\ 
    Stuttgart 70599, GERMANY \\
    Tel.: (+49) 0711 459-22334 \\
    Fax: (+49) 0711 459-22343 \\ 
    \texttt{melchinger@uni-hohenheim.de} \vfill}

@@

% Introduction
\section{Introduction} 
% Rationale for predictions.
Genomic prediction, pioneered by \citeNP{Meuwissen2001}, has revolutionized
animal and plant breeding by a more efficient selection of promising candidates 
without phenotypes \cite{DeLosCampos2013,Garcia-Ruiz2016} and recently emerged
as an important tool in human diagnostics \cite{DeLosCampos2010,Vazquez2016}.
Its use is largely motivated by the high costs and oftentimes long duration of 
phenotyping, which can be reduced when covering all genotypes with a predictor 
while limiting phenotyping to a few genotypes with high precision
\cite{Kadam2016}.
The performance of all other genotypes is then forecast using model parameters
obtained from thorough resampling routines applied to the subset of individuals 
with full information \cite{Hastie2009}.
In animal breeding, genomic selection has increased the annual selection gain 
by $\approx$ 50-100\% for yield traits and by 300-400\% for traits with low 
heritability \cite{Garcia-Ruiz2016}.

% Incomplete predictors.
Breeding programs --- particularly in animal breeding --- are characterized by
the presence of one predictor type, which is available for all recorded
genotypes while other predictors are incomplete in that they cover only a 
subset of all genotypes \cite{Fragomeni2015}.
Usually, the complete predictor is pedigree information while genomic data are
the most prevalent incomplete predictor.
To utilize all available phenotypic information for model training, methods were 
developed that allowed for a combined use of genomic and pedigree information 
\cite{Hayes2009a,VanRaden2009}.

% Two-step prediction.
Initially, a two-step procedure was the most intuitive choice for accomplishing
this goal \cite{VanRaden2009}.
The first step consists of conventionally estimating breeding values based on
pedigree relationships and the phenotypes of relatives.
Individuals with accurate breeding values out of the first step (\textit{i.e.}
high number of offspring), enter the second step in which conventional breeding
values are regressed onto SNP genotypes.
The estimated model coefficients can then be used to predict genomic breeding 
values for solely genotyped and potentially very young individuals.
However, as the breeding values from step one enter the genomic prediction
model as phenotypes, one has to be careful about the residual error
distribution of such a model.
For instance, the range of the prediction errors will no longer be identically
distributed, \textit{i.e.}, they should be weighted differently or estimated 
individually \cite{Aguilar2010}.
One popular approach to account for this problem is deregressing the breeding
values and weighting the residuals according to the prediction error variances
\cite{Garrick2009}.

% Workings and advantages of the single step procedure.
Independently from each other, \citeNP{Legarra2009} and
\citeNP{Christensen2010} developed a single step BLUP method that blends the 
numerator relationship matrix $\mathbf{A}$ with the genomic relationship matrix
$\mathbf{G}$ in a mutual matrix $\mathbf{H}$ to use all predictor information
simultaneously.
Issues with this H-BLUP approach were the compatability of $\mathbf{A}$ with
$\mathbf{G}$ \cite{Christensen2012} and their weighting, which is not
trait-independent \cite{Vitezica2011,Ashraf2016}.
\citeNP{Fernando2014} derived the equivalent single step marker-effect-model to 
the single step breeding value model of \citeNP{Legarra2009} and 
\citeNP{Christensen2010}, which does not require weights for $\mathbf{A}$ and 
$\mathbf{G}$.
An additional advantage of this alternative formulation is the ability to 
implicitly model the structure of the imputation error.

% -   Mention scarcity of data on single step prediction in plant populations.
In animal breeding, single step prediction, particularly H-BLUP, has already 
become a routine procedure \cite{Legarra2014} but, to the best of our knowledge, 
has only been considered in plants in a study on wheat \cite{Ashraf2016} and
white spruce lines, respectively \cite{Ratcliffe2017}. 
Our objectives were threefold: (1) Examine whether the single step prediction
framework can be transfered to plant hybrids, (2) explore the utility of single 
step predition with a quantitative predictor and (3) evaluate the impact of the
genetic space covered by the incomplete predictor on predictive ability.

\section{Materials and Methods}
\Genetics2level{Maize hybrids}
The maize hybrid data set, subsequently denoted as \textit{Exp1}, comprised 
1,521 hybrids produced in 16 factorial mating designs between 142 Dent and 103 
Flint parent lines \cite{Westhues2017}. 
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse locations across Germany for six 
agronomically important traits by accounting for year, location, field 
replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper \cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50 \cite{Ganal2011}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
using the Beagle software \cite{Browning2009}.
The final number of polymorphic marker loci was 7,013 for the Dent and 6,212 for 
the Flint lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 parental Dent and 43 parental Flint lines, that are parents of 685
hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267) \cite{Westhues2017}.
After applying established normalization procedures \cite{Smyth2003,Ritchie2007},
gene expression BLUEs were computed for 1,323 transcripts while adjusting for 
putative effects arising from the presence of multiple factorials (\textit{cf.}
\citeNP{Westhues2017}) using the R-package \textit{limma} \cite{Ritchie2015a}.

@@

\Genetics2level{Maize diversity panel}
The maize inbred line data set, subsequently referred to as \textit{Exp2}, 
originally comprised 513 maize lines representing the global maize diversity 
and was reduced to the set of tropical and subtropical lines ($n = 211$) given 
that it is the largest of the four pre-classified subgroups \cite{Yang2014}.
We reduced the original dataset to this subset to avoid inflated predictive 
abilities due to large differences in the average performance of the different
subpopulations \cite{Windhausen2012}.
All inbred lines were evaluated in five Chinese environments ranging from
$18$ to $30^{\circ}$N and from $102$ to $110^{\circ}$E \cite{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated for 17 traits of which 
six are presented in this study (Table~\ref{table:TraitKey}).
The selection of these traits was done so that four traits were better 
predicted by transcriptomic data and two further traits were better predicted 
by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploting identify by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines \cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers \cite{Yang2014}.
The same SNP quality checks as for the parental inbred lines of the hybrids
were applied to the 211 inbred lines from the tropical/subtropical subset,
yielding 37,760 SNPs.
Gene expression data were normalized using a normal quantile transformation to
satisfy modelling assumptions \cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes and was kept for further analyses \cite{Li2013}.

@@

\Genetics2level{Population structure}
Separately for each data set, principal component analyses (PCA) were run on 
scaled and centered predictor matrices using the \emph{pca} function from the 
R-package \textit{LEA} \cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms \cite{Frichot2014} using the 
\emph{snmf} function in \textit{LEA}.
Algorithms were run for $K=2, \dots 5$ putative ancestral gene pools with 25
repetitions for each value of $K$. 
For each $K$, only the repetition with the lowest cross-entropy was kept for 
further analyses.

\Genetics2level{Core sampling}
In order to evaluate the influence of the genetic constitution of the set of
genotypes that has only data on one out of two predictors, we generated core
samples based on two different approaches.
All core sets were assembled from the 149 maize inbred lines from the diversity
panel which had gene expression as well as genomic information.
We defined core samples as a subset of genotypes that is covered by two
predictors whereas all other genotypes (\textit{i.e.}, the complement) are only
covered by a single predictor.
The first application involved assembling core sets of varying sizes by 
maximizing the average genetic distance among core set members using the 
Modified Rogers (MR) distance as the criterion \cite{Thachuk2009}.
The size of the core set was varied in increments of ten percentage points and 
ranged from 10\% to 90\% of all 149 inbred lines.
For the second application, we first ran an admixture analysis in combination
with a PCA and kept $K=3$ as the number of putative ancestral population, which 
seemed to describe the population structure best.
Genotypes were then assigned to one of the three ancestral populations based on
whether any of the three relationship coefficients (A1-A3) was equal to or
greater than 0.5.
For example, a genotype with an admixture coefficient for A2 $\geq 0.5$ was
assigned to the second ancestral population.
Thereby, we obtained $n_{A1} = 17$, $n_{A2} = 109$, $n_{A3} = 19$ and four 
genotypes that could not be unambiguously assigned to any putative ancestral 
population based on this criterion.
To ensure commensurability of the three subpopulations within the set of 149
genotypes, the subset of 109 genotypes (\textit{i.e.}, members of A2) was
reduced to $n_{A2} = 19$ genotypes, again using the method of maximizing the
average genetic distance among selected genotypes described above.

@@

where $W_{l}$ denotes the number of features \cite{VanRaden2008}.
In the case of pedigree data (P), coancestry coefficients were used directly
for $\mathbf{K}_{l}$.

@@

where $\mathbf{y}$ is the vector of observed inbred line or hybrid performance,
respectively, $\mu$ is the fixed model intercept, and $\mathbf{e}$ is the 
residual error.
For the hybrid data from \textit{Exp1}, we fit separate additive effects for 
each of the $L$ heterotic groups and $\mathbf{g}_{l}$ now pertains to the 
general combining ability effects of the parental inbred lines (either Dent or 
Flint, as expressed by $l$).
For the maize diversity panel inbred lines from \textit{Exp2}, we fit random 
genotype effects $\mathbf{g}_{l}$ (\textit{i.e.}, breeding values) of the 
lines for the $o$-th predictor.
For both experiments, genetic effects $\mathbf{g}_{l}$ are associated with 
the vector of phenotypic performance $\mathbf{y}$ through a corresponding 
design matrix $\mathbf{Z}_{l}$.
The random effects ($\mathbf{g}_{l}$) have expectation zero and variance equal 
to $\mathbf{K}_{l} \sigma^{2}_{{g}^{l}}$ and $\mathbf{I} \sigma^2_{e}$ 
for the residual error.
Here, we did not consider the inclusion of dominance effects for the hybrid
material after a previous investigation into the predictive ability of various
"omics" predictors in the same material concluded no considerably gains 
\cite{Westhues2017}.
All prediction models were implemented using the R-packages \textit{BGLR}
\cite{Perez2014} and \textit{sspredr}
\url{https://github.com/mwesthues/sspredr}.

@@

\begin{align} \label{eq:KBLUPModelSimplified}
	\mathbf{y} &= \mu + \mathbf{Z} \mathbf{g} + \boldsymbol{e} \\
	&= \mu + \mathbf{Z} \mathbf{W} \boldsymbol{\alpha} + \boldsymbol{e}
\end{align}
\cite{Fernando2014} since the ridge regression model is equivalent to the BLUP
model \cite{Ruppert2003}.

@@

where $\mathbf{K}_{ij}$ are partitions of $\mathbf{K}$ that correspond to 
$\mathbf{g}_{1}$ and $\mathbf{g}_{2}$, respectively \cite{Fernando2014}.

@@

\section*{Results}
\Genetics2level{Inbred lines}
At the level of the 149 genotypes for which both, genomic (G) and 
transcriptomic (T) information, were available, the use of T yielded higher
predictive abilities than the use of G for four out of six traits
(Fig. \ref{fig:InbredResults}A).
In terms of predictive ability, the absolute advantage of T over G was as large 
as 0.04 as observed for the traits DS, ED, GW and KW.
Except for KW, the predictive ability for predictions based on G were always
higher for the full set of inbred lines ($n = 211$) than for the reduced set of 
inbred lines ($n = 149$).
The combination of G and T (\textit{i.e.}, GT) in a single-step approach for the 
prediction of all 211 lines, performed at least as well as G alone when T was 
superior over G for the reduced set of 149 lines.
In the case of GW and ED, the use of GT yielded a slight improvement in predictive 
ability over G.
Where T was a worse predictor for the reduced set of 149 lines than G,
predictive abilities of GT for the full set of 211 lines were also worse than 
those based on G alone.

% Script: ./analysis/maizego_core_fraction_plot_predictive_abilities.R
\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/inbred_line_combi_plot.pdf}
  \caption{
  Predictive abilities and average coefficient of variation for
  tropical/subtropical inbred lines from the maize diversity panel and six 
  agronomic traits.
  As predictors, genomic (G) data, transcriptomic (T) data and their 
  combination (GT) were used.
  (\textbf{A}) The 'Core' panel includes a subset of 149 lines covered by
  genomic and transcriptomic information whereas the 'Full' panel comprises all
  211 genotypes. In the 'Full' set of genotypes, transcriptomic records were
  imputed for 62 lines.
  (\textbf{B}) Influence of structural missingness in the coverage of genotypes 
  with transcriptomic information on single step-based predictive abilities 
  using a 'Core' set of 149 inbred lines.
  Transcriptomic records were deleted for 17 to 19 genotypes that were
  primarily associated with a particular subgroup in the material, as defined
  by an admixture analysis and a PCA (Fig. \ref{fig:PopStructure}).
  (\textbf{C}) Influence of core set sampling on predictive abilities using
  nine different sub set sizes where 'Core Fraction' indicates the fraction
  of 149 inbred lines that is covered by both, genomic and transcriptomic
  data.
  } 
  \label{fig:InbredResults}
\end{figure}

\Genetics2level{Hybrids}
The maize hybrid data set contains information on six agronomic traits and
three predictors, namely pedigree data (P), genomic information (G), 
transcriptomic information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the 'Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as 'Core'.
For DMY, T was the best predictor in the core set of 685 hybrids with a 
predictive ability that was 0.08 points higher than for genomic data, as the 
second best predictor for this trait (Fig. \ref{fig:HybridResults}).
In the core set, P had the highest predictive ability for STA and DMC.
For the other three traits, G was the superior predictor inside the core set.
In the full set of 1,521 hybrids, predictive abilities obtained with G rose 
markedly for all traits with the exception of FAT.
The combination of genomic and transriptomic information (GT) in a single-step
model yielded a small improvement over G alone for the traits DMY and PRO.
Except for PRO, compared to using only G, trait prediction did not benefit from 
the single step approach when T was not superior to G in the core set.
Predictive abilities resulting from combinations of pedigree with genomic (PG)
information were better than P alone for all traits and PT improved upon P in
the core set for four traits.

% Script: ./analysis/uhoh_plot_predictive_abilities.R
\begin{figure}[H]
\centering
  \includegraphics{./tables_figures/hybrid_predictive_ability.pdf}
  \caption{
    Predictive abilities and average coefficients of variation for the set of 
    maize hybrids and six agronomic traits.
    As predictors, pedigree (P), genomic (G), transcriptomic (T) data and 
    combinations thereof we used.
    The 'Full' set of genotypes comprises 1,521 hybrid progeny based on 245
    parental inbred lines, whereas the 'Core' set comprises 685 hybrid progeny
    based on 103 parental inbred lines.
  } 
  \label{fig:HybridResults}
\end{figure}

\Genetics2level{Coverage of the genetic space}
Single-step approaches are based on the use of one complete predictor that
covers all genotypes and one incomplete predictor that covers only a subset of
all available genotypes.
Here, we explore the impact of which data are covered by the incomplete 
predictor on the predictive abilities obtained via single-step approaches by
focusing exclusively on the 149 inbred lines from the maize diversity panel
that are covered by both, genomic and transcriptomic information.
First, we assembled nine different core sets, which we defined as subsets
of genotypes that maximize the average genetic distance between genotypes of
the full set.
The nine core sets differ in that each covers a different number of genotypes,
ranging from only 10\% to 90\% of all genotypes covered by the incomplete
predictor (Fig. \ref{fig:CoreSetPCA}).

@@

Predictive abilities obtained with the different core sets varied widely within
traits (Fig. \ref{fig:InbredResults}B).
For the traits CW and PH, which were better predicted via G than via T in the 
reduced set of lines, predictive abilities were higher when only a small number 
of genotypes was included in the core set.
This corresponds to a situation where a large number of genotypes is being 
imputed and the majority of the information is derived from genomic information. 
For most other traits, an initial increase in predictive abilities when moving
from a core set size of 0.1 to 0.5/0.6 was observed, followed by a drop around
0.6/0.7 and a subsequent hike up to a core set size of 0.9.
A principal component analysis depicts the coverage of the genetic space by
genotypes (Fig. \ref{fig:PopStructure}).
With a core set size of 50-60\% of the 149 genotypes, the congruence between
core set and complement was high whereas genotypes in the south-west direction
of the PCA plot were overrepresented compared to genotypes in the east 
direction at a core set size of 0.7.

>>>>

\renewcommand{\CorrespondingAddress}{%
    University of Hohenheim \\
    Institute of Plant Breeding, Seed Science and Population Genetics \\ 
    Fruwirthstr. 21 \\ 
    Stuttgart 70599, GERMANY \\
    Tel.: (+49) 0711 459--22334 \\
    Fax: (+49) 0711 459--22343 \\ 
    \texttt{melchinger@uni-hohenheim.de} \vfill}

@@

% Introduction
\section{Introduction}
% Rationale for predictions.
Genomic prediction, pioneered by \citeNP{Meuwissen2001}, has revolutionized
animal and plant breeding by a more efficient selection of promising candidates 
without phenotypes~\cite{DeLosCampos2013,Garcia-Ruiz2016} and recently emerged
as an important tool in human diagnostics~\cite{DeLosCampos2010,Vazquez2016}.
Its use is largely motivated by the high costs and oftentimes long duration of 
phenotyping, which can be reduced when covering all genotypes with a predictor 
while limiting phenotyping to a few genotypes with high precision
\cite{Kadam2016}.
The performance of all other genotypes is then forecast using model parameters
obtained from thorough resampling routines applied to the subset of individuals 
with full information~\cite{Hastie2009}.
In animal breeding, genomic selection has increased the annual selection gain 
by $\approx$ 50--100\% for yield traits and by 300--400\% for traits with low 
heritability~\cite{Garcia-Ruiz2016}.

% Incomplete predictors.
Breeding programs --- particularly in animal breeding --- are characterized by
the presence of one predictor type, which is available for all recorded
genotypes while other predictors are incomplete in that they cover only a 
subset of all genotypes~\cite{Fragomeni2015}.
Usually, the complete predictor is pedigree information while genomic data are
the most prevalent incomplete predictor.
To utilize all available phenotypic information for model training, methods were 
developed that allowed for a combined use of genomic and pedigree information 
\cite{Hayes2009a,VanRaden2009}.

% Two-step prediction.
Initially, a two-step procedure was the most intuitive choice for accomplishing
this goal~\cite{VanRaden2009}.
The first step consists of conventionally estimating breeding values based on
pedigree relationships and the phenotypes of relatives.
Individuals with accurate breeding values out of the first step (that is a
high number of offspring), enter the second step in which conventional breeding
values are regressed onto SNP genotypes.
The estimated model coefficients can then be used to predict genomic breeding 
values for solely genotyped and potentially very young individuals.
However, as the breeding values from step one enter the genomic prediction
model as phenotypes, one has to be careful about the residual error
distribution of such a model.
For instance, the range of the prediction errors will no longer be identically
distributed, that is they should be weighted differently or estimated 
individually~\cite{Aguilar2010}.
One popular approach to account for this problem is deregressing the breeding
values and weighting the residuals according to the prediction error variances
\cite{Garrick2009}.

% Workings and advantages of the single step procedure.
Independently from each other,~\citeNP{Legarra2009} and
\citeNP{Christensen2010} developed a single step BLUP method that blends the 
numerator relationship matrix $\mathbf{A}$ with the genomic relationship matrix
$\mathbf{G}$ in a mutual matrix $\mathbf{H}$ to use all predictor information
simultaneously.
Issues with this H-BLUP approach were the compatability of $\mathbf{A}$ with
$\mathbf{G}$~\cite{Christensen2012} and their weighting, which is not
trait-independent~\cite{Vitezica2011,Ashraf2016}.
\citeNP{Fernando2014} derived the equivalent single step marker-effect-model to 
the single step breeding value model of \citeNP{Legarra2009} and 
\citeNP{Christensen2010}, which does not require weights for $\mathbf{A}$ and 
$\mathbf{G}$ and allows implicit modeling of the imputation error's structure.

% -   Mention scarcity of data on single step prediction in plant populations.
In animal breeding, single step prediction, particularly H-BLUP, has already 
become a routine procedure~\cite{Legarra2014} but, to the best of our knowledge, 
has only been considered in plants in a study on wheat~\cite{Ashraf2016} and
white spruce lines, respectively~\cite{Ratcliffe2017}.
Our objectives were threefold: (1) Examine whether the single step prediction
framework can be transfered to plant hybrids, (2) explore the utility of single 
step predition with a quantitative predictor and (3) evaluate the impact of the
proporition of genotypes not covered by the incomplete predictor on predictive
ability.

\section{Materials and Methods}
\Genetics2level{Maize hybrids}
The maize hybrid data set, subsequently denoted as \textit{Exp1}, comprised 
1,521 hybrids produced in 16 factorial mating designs between 142 Dent and 103 
Flint parent lines~\cite{Westhues2017}. 
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse locations across Germany for six 
agronomically important traits by accounting for year, location, field 
replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
using the Beagle software~\cite{Browning2009}.
The final number of polymorphic marker loci was 7,013 for the Dent and 6,212 for 
the Flint lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 parental Dent and 43 parental Flint lines, that are parents of 685
hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
gene expression BLUEs were computed for 1,323 transcripts while adjusting for 
putative effects arising from the presence of multiple factorials (\textit{cf.}
\citeNP{Westhues2017}) using the R-package \textit{limma}~\cite{Ritchie2015a}.

@@

\Genetics2level{Maize diversity panel}
The maize inbred line data set, subsequently referred to as \textit{Exp2}, 
originally comprised 513 maize lines representing the global maize diversity 
and was reduced to the set of tropical and subtropical lines ($n = 211$) given 
that it is the largest of the four pre-classified subgroups~\cite{Yang2014}.
We reduced the original dataset to this subset to avoid inflated predictive 
abilities due to large differences in the average performance of the different
subpopulations~\cite{Windhausen2012}.
All inbred lines were evaluated in five Chinese environments ranging from
$18$ to $30^{\circ}$N and from $102$ to $110^{\circ}$E~\cite{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated for 17 traits of which 
six are presented in this study (Table~\ref{table:TraitKey}).
The selection of these traits was done so that four traits were better 
predicted by transcriptomic data and two further traits were better predicted 
by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploting identify by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines~\cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers~\cite{Yang2014}.
The same SNP quality checks as for the parental inbred lines of the hybrids
were applied to the 211 inbred lines from the tropical/subtropical subset,
yielding 37,760 SNPs.
Gene expression data were normalized using a normal quantile transformation to
satisfy modelling assumptions~\cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes and was kept for further analyses~\cite{Li2013}.

@@

\Genetics2level{Population structure}
Separately for each data set, principal component analyses (PCA) were run on 
scaled and centered predictor matrices using the \emph{pca} function from the 
R-package \textit{LEA}~\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \textit{LEA}.
Algorithms were run for $K=2, \dots 5$ putative ancestral gene pools with 25
repetitions for each value of $K$. 
For each $K$, only the repetition with the lowest cross-entropy was kept for 
further analyses.

\Genetics2level{Core sampling}
In order to evaluate the influence of the genetic constitution of the set of
genotypes that has only data on one out of two predictors, we generated core
samples based on two different approaches.
All core sets were assembled from the 149 maize inbred lines from the diversity
panel which had gene expression as well as genomic information.
We defined core samples as a subset of genotypes that is covered by two
predictors whereas all other genotypes (\textit{i.e.}, the complement) are only
covered by a single predictor.
The first application involved assembling core sets of varying sizes by 
maximizing the average genetic distance among core set members using the 
Modified Rogers (MR) distance as the criterion~\cite{Thachuk2009}.
The size of the core set was varied in increments of ten percentage points and 
ranged from 10\% to 90\% of all 149 inbred lines.
For the second application, we first ran an admixture analysis in combination
with a PCA and kept $K=3$ as the number of putative ancestral population, which 
seemed to describe the population structure best.
Genotypes were then assigned to one of the three ancestral populations based on
whether any of the three relationship coefficients (A1-A3) was equal to or
greater than 0.5.
For example, a genotype with an admixture coefficient for A2 $\geq 0.5$ was
assigned to the second ancestral population.
Thereby, we obtained $n_{A1} = 17$, $n_{A2} = 109$, $n_{A3} = 19$ and four 
genotypes that could not be unambiguously assigned to any putative ancestral 
population based on this criterion.
To ensure commensurability of the three subpopulations within the set of 149
genotypes, the subset of 109 genotypes (\textit{i.e.}, members of A2) was
reduced to $n_{A2} = 19$ genotypes, again using the method of maximizing the
average genetic distance among selected genotypes described above.

@@

where $W_{l}$ denotes the number of features~\cite{VanRaden2008}.
In the case of pedigree data (P), coancestry coefficients were used directly
for $\mathbf{K}_{l}$.

@@

where $\mathbf{y}$ is the vector of observed inbred line or hybrid performance,
respectively, $\mu$ is the fixed model intercept, and $\mathbf{e}$ is the 
residual error.
For the hybrid data from \textit{Exp1}, we fit separate additive effects for 
each of the $L$ heterotic groups and $\mathbf{g}_{l}$ now pertains to the 
general combining ability effects of the parental inbred lines (either Dent or 
Flint, as expressed by $l$).
For the maize diversity panel inbred lines from \textit{Exp2}, we fit random 
genotype effects $\mathbf{g}_{l}$ (\textit{i.e.}, breeding values) of the 
lines for the $o$-th predictor.
For both experiments, genetic effects $\mathbf{g}_{l}$ are associated with 
the vector of phenotypic performance $\mathbf{y}$ through a corresponding 
design matrix $\mathbf{Z}_{l}$.
The random effects ($\mathbf{g}_{l}$) have expectation zero and variance equal 
to $\mathbf{K}_{l} \sigma^{2}_{{g}^{l}}$ and $\mathbf{I} \sigma^2_{e}$ 
for the residual error.
Here, we did not consider the inclusion of dominance effects for the hybrid
material after a previous investigation into the predictive ability of various
``omics'' predictors in the same material concluded no considerably gains 
\cite{Westhues2017}.
All prediction models were implemented using the R-packages \textit{BGLR}
\cite{Perez2014} and \textit{sspredr}
\url{https://github.com/mwesthues/sspredr}.

@@

\begin{align} \label{eq:KBLUPModelSimplified}
	\mathbf{y} &= \mu + \mathbf{Z} \mathbf{g} + \boldsymbol{e} \\
	&= \mu + \mathbf{Z} \mathbf{W} \boldsymbol{\alpha} + \boldsymbol{e}
\end{align}
\cite{Fernando2014} since the ridge regression model is equivalent to the BLUP
model~\cite{Ruppert2003}.

@@

where $\mathbf{K}_{ij}$ are partitions of $\mathbf{K}$ that correspond to 
$\mathbf{g}_{1}$ and $\mathbf{g}_{2}$, respectively~\cite{Fernando2014}.

@@

\section*{Results}
\Genetics2level{Inbred lines}
At the level of the 149 genotypes for which both, genomic (G) and 
transcriptomic (T) information, were available, the use of T yielded higher
predictive abilities than the use of G for four out of six traits
(Fig.~\ref{fig:InbredResults}A).
In terms of predictive ability, the absolute advantage of T over G was as large 
as 0.04 as observed for the traits DS, ED, GW and KW\@.
Except for KW, the predictive ability for predictions based on G were always
higher for the full set of inbred lines ($n = 211$) than for the reduced set of 
inbred lines ($n = 149$).
The combination of G and T (\textit{i.e.}, GT) in a single-step approach for the 
prediction of all 211 lines, performed at least as well as G alone when T was 
superior over G for the reduced set of 149 lines.
In the case of GW and ED, the use of GT yielded a slight improvement in predictive 
ability over G.
Where T was a worse predictor for the reduced set of 149 lines than G,
predictive abilities of GT for the full set of 211 lines were also worse than 
those based on G alone.

% Script: ./analysis/maizego_core_fraction_plot_predictive_abilities.R
\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/inbred_line_combi_plot.pdf}
  \caption{
  Predictive abilities and average coefficient of variation for
  tropical/subtropical inbred lines from the maize diversity panel and six 
  agronomic traits.
  As predictors, genomic (G) data, transcriptomic (T) data and their 
  combination (GT) were used.
  (\textbf{A}) The `Core' panel includes a subset of 149 lines covered by
  genomic and transcriptomic information whereas the `Full' panel comprises all
  211 genotypes. In the `Full' set of genotypes, transcriptomic records were
  imputed for 62 lines.
  (\textbf{B}) Influence of structural missingness in the coverage of genotypes 
  with transcriptomic information on single step-based predictive abilities 
  using a `Core' set of 149 inbred lines.
  Transcriptomic records were deleted for 17 to 19 genotypes that were
  primarily associated with a particular subgroup in the material, as defined
  by an admixture analysis and a PCA (Fig.~\ref{fig:PopStructure}).
  (\textbf{C}) Influence of core set sampling on predictive abilities using
  nine different sub set sizes where `Core Fraction' indicates the fraction
  of 149 inbred lines that is covered by both, genomic and transcriptomic
  data.
  }
\label{fig:InbredResults}
\end{figure}

\Genetics2level{Hybrids}
The maize hybrid data set contains information on six agronomic traits and
three predictors, namely pedigree data (P), genomic information (G), 
transcriptomic information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the `Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as `Core'.
For DMY, T was the best predictor in the core set of 685 hybrids with a 
predictive ability that was 0.08 points higher than for genomic data, as the 
second best predictor for this trait (Fig.~\ref{fig:HybridResults}).
In the core set, P had the highest predictive ability for STA and DMC\@.
For the other three traits, G was the superior predictor inside the core set.
In the full set of 1,521 hybrids, predictive abilities obtained with G rose 
markedly for all traits with the exception of FAT\@.
The combination of genomic and transriptomic information (GT) in a single-step
model yielded a small improvement over G alone for the traits DMY and PRO\@.
Except for PRO, compared to using only G, trait prediction did not benefit from 
the single step approach when T was not superior to G in the core set.
Predictive abilities resulting from combinations of pedigree with genomic (PG)
information were better than P alone for all traits and PT improved upon P in
the core set for four traits.

% Script: ./analysis/uhoh_plot_predictive_abilities.R
\begin{figure}[H]
\centering
  \includegraphics{./tables_figures/hybrid_predictive_ability.pdf}
  \caption{
    Predictive abilities and average coefficients of variation for the set of 
    maize hybrids and six agronomic traits.
    As predictors, pedigree (P), genomic (G), transcriptomic (T) data and 
    combinations thereof we used.
    The `Full' set of genotypes comprises 1,521 hybrid progeny based on 245
    parental inbred lines, whereas the `Core' set comprises 685 hybrid progeny
    based on 103 parental inbred lines.
  }
\label{fig:HybridResults}
\end{figure}

\Genetics2level{Coverage of the genetic space}
Single-step approaches are based on the use of one complete predictor that
covers all genotypes and one incomplete predictor that covers only a subset of
all available genotypes.
Here, we explore the impact of which data are covered by the incomplete 
predictor on the predictive abilities obtained via single-step approaches by
focusing exclusively on the 149 inbred lines from the maize diversity panel
that are covered by both, genomic and transcriptomic information.
First, we assembled nine different core sets, which we defined as subsets
of genotypes that maximize the average genetic distance between genotypes of
the full set.
The nine core sets differ in that each covers a different number of genotypes,
ranging from only 10\% to 90\% of all genotypes covered by the incomplete
predictor (Fig.~\ref{fig:CoreSetPCA}).

@@

Predictive abilities obtained with the different core sets varied widely within
traits (Fig.~\ref{fig:InbredResults}B).
For the traits CW and PH, which were better predicted via G than via T in the 
reduced set of lines, predictive abilities were higher when only a small number 
of genotypes was included in the core set.
This corresponds to a situation where a large number of genotypes is being 
imputed and the majority of the information is derived from genomic information. 
For most other traits, an initial increase in predictive abilities when moving
from a core set size of 0.1 to 0.5/0.6 was observed, followed by a drop around
0.6/0.7 and a subsequent hike up to a core set size of 0.9.
A principal component analysis depicts the coverage of the genetic space by
genotypes (Fig.~\ref{fig:PopStructure}).
With a core set size of 50--60\% of the 149 genotypes, the congruence between
core set and complement was high whereas genotypes in the south-west direction
of the PCA plot were overrepresented compared to genotypes in the east 
direction at a core set size of 0.7.

As an alternative to exploring the impact of genotypes sampled based on the
maximization of genetic distance within the core set, we wanted to explore a
more extreme scenario where we select core sets based on population structure.
Hereto, we ran an admixture analysis on the same set of 149 inbred lines 
assuming $K=3$ ancestral populations (Fig.~\ref{fig:PopStructure}).
